---
title: "Disease alleles, penetrance and population risk"
subtitle: "Population Genetics Practical 13: From genotype frequencies to phenotype probabilities"
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show R code"
    embed-resources: true
editor: source
execute:
  error: false
---

## Learning objectives

Distinguish allele frequency, heterozygote frequency and phenotype probability; apply genotype-specific penetrance; examine an inbreeding model; distinguish population averages from known-parent crosses; and identify why phenotype prevalence alone may not identify an allele frequency.

This practical follows [population genetics in medicine](../slides/introduction_population_genetics.qmd#population-genetics-in-medicine), building on [HWE](allele_frequencies_hwe.qmd) and [structure and inbreeding](structure_mating_inbreeding.qmd). Every parameter is hypothetical and describes an invented single-locus phenotype. The examples use base R and are not estimates for a named disease or an individual.

## 1. Define the allele and the observation window

Let a have frequency $q$ and A frequency $p=1-q$ at an autosomal diploid locus. Under random union, genotype probabilities are $p^2,2pq,q^2$ for AA, Aa and aa.

For this example, call Aa a *heterozygous carrier* of a. An aa individual has two copies and is reported separately. Keep the counted allele and genotype labels explicit.

```{r}
#| label: genotype-frequencies
q <- 0.02
genotype_frequencies <- function(q,F=0) {
  stopifnot(length(q)==1,is.finite(q),q>=0,q<=1,
            length(F)==1,is.finite(F),F>=0,F<=1)
  p <- 1-q
  c(AA=p^2+F*p*q,Aa=2*p*q*(1-F),aa=q^2+F*p*q)
}
freq <- genotype_frequencies(q)
stopifnot(abs(sum(freq)-1)<1e-12,
          max(abs(freq-c(0.9604,0.0392,0.0004)))<1e-12)
data.frame(genotype=names(freq),probability=unname(freq),
           expected_in_100000=100000*unname(freq))
```

The expected counts are averages over hypothetical samples; they do not require every sample to contain exactly these numbers.

Define $f_g=P(Y=1\mid G=g)$ as the probability of the specified phenotype by a declared age or observation window. These genotype-specific probabilities are the penetrance inputs. They describe whether the phenotype occurs, rather than its severity. MedlinePlus explains [reduced penetrance and variable expressivity](https://medlineplus.gov/genetics/understanding/inheritance/penetranceexpressivity/).

## 2. Average genotype-specific phenotype probabilities

The population phenotype probability is

$$K=P(Y=1)=\sum_g P(G=g)f_g.$$

In a fully penetrant, purely recessive model, $f_{AA}=f_{Aa}=0$ and $f_{aa}=1$. Only with these assumptions and HWE does $K=q^2$.

```{r}
#| label: penetrance-models
full_recessive <- c(AA=0,Aa=0,aa=1)
incomplete <- c(AA=0,Aa=0,aa=0.8)
background <- c(AA=0.001,Aa=0.001,aa=0.8)
phenotype_probability <- function(freq,penetrance) {
  stopifnot(identical(names(freq),names(penetrance)),
            all(is.finite(freq)),all(freq>=0),abs(sum(freq)-1)<1e-12,
            all(is.finite(penetrance)),all(penetrance>=0),all(penetrance<=1))
  sum(freq*penetrance)
}
models <- list(Full_recessive=full_recessive,Incomplete=incomplete,
               Background=background)
K <- vapply(models,function(f) phenotype_probability(freq,f),numeric(1))
stopifnot(max(abs(K-c(0.0004,0.00032,0.0013196)))<1e-12)
K
```

The background model allows the same phenotype in people without aa through causes outside this locus. It does not classify those causes or assign a clinical diagnosis. Its parameters must refer to the same observation window and sampling population.

**Question.** Does an aa genotype guarantee the phenotype in the incomplete model?

::: {.callout-note collapse="true" title="Suggested answer"}
No. Its supplied phenotype probability is 0.8. Conversely, in the background model the phenotype can occur without aa. Genotype, phenotype and inferred cause are distinct.
:::

## 3. Inbreeding changes genotype frequencies

Under the inbreeding model on a specified ancestral base,

$$P(AA)=p^2+Fpq,\quad P(Aa)=2pq(1-F),\quad P(aa)=q^2+Fpq.$$

This preserves allele frequency while shifting heterozygote probability into both homozygote classes. Here $F$ is a supplied model parameter, not inferred from phenotype prevalence or a heterozygosity deficit.

```{r}
#| label: inbreeding-comparison
F_values <- c(0,0.01,0.0625)
comparison <- do.call(rbind,lapply(F_values,function(F) {
  g <- genotype_frequencies(q,F)
  stopifnot(abs(g["aa"]+g["Aa"]/2-q)<1e-12)
  data.frame(F=F,heterozygote=unname(g["Aa"]),aa=unname(g["aa"]),
             phenotype=phenotype_probability(g,background))
}))
comparison
```

```{r}
#| label: inbreeding-risk-plot
F_grid <- seq(0,0.1,length.out=100)
plot(F_grid,vapply(F_grid,function(F) {
  phenotype_probability(genotype_frequencies(q,F),background)
},numeric(1)),type="l",lwd=2,col="#1b9e77",
     xlab="Supplied inbreeding coefficient F",ylab="Population phenotype probability")
```

For this particular penetrance vector, the phenotype probability increases with F. More generally,

$$K(F)-K(0)=Fpq(f_{AA}-2f_{Aa}+f_{aa}).$$

The direction depends on genotype-specific phenotype probabilities; it is not universally positive. This one-generation model does not simulate selection, purging, demography or changes in q.

```{r}
#| label: inbreeding-identity-check
g_F <- genotype_frequencies(q,0.0625)
change <- phenotype_probability(g_F,background)-K["Background"]
expected_change <- 0.0625*q*(1-q)*
  (background["AA"]-2*background["Aa"]+background["aa"])
stopifnot(abs(change-expected_change)<1e-12)
```

## 4. Condition on known parents

A population-average probability does not replace a probability conditional on known parental genotypes. For Aa by Aa, each parent transmits a with probability 1/2. Mendelian transmission gives AA, Aa and aa probabilities 1/4, 1/2 and 1/4.

```{r}
#| label: known-parent-crosses
cross_probabilities <- function(sire_dosage,dam_dosage) {
  stopifnot(length(sire_dosage)==1,length(dam_dosage)==1,
            sire_dosage %in% 0:2,dam_dosage %in% 0:2)
  ps <- sire_dosage/2
  pd <- dam_dosage/2
  c(AA=(1-ps)*(1-pd),Aa=ps*(1-pd)+(1-ps)*pd,aa=ps*pd)
}
carrier_cross <- cross_probabilities(1,1)
stopifnot(max(abs(carrier_cross-c(0.25,0.5,0.25)))<1e-12)
data.frame(genotype=names(carrier_cross),probability=unname(carrier_cross))
c(full_recessive=phenotype_probability(carrier_cross,full_recessive),
  incomplete=phenotype_probability(carrier_cross,incomplete),
  background=phenotype_probability(carrier_cross,background))
```

For this cross, aa probability is 0.25 while phenotype probability in the incomplete model is 0.20. These are per-offspring probabilities under the declared model. The cross calculation needs the parental genotypes rather than the population q; no HWE assumption is needed once those genotypes are known.

## 5. Reverse conditioning is a different calculation

In the background population model, Bayes' rule gives

$$P(G=g\mid Y=1)=\frac{P(G=g)f_g}{K}.$$

```{r}
#| label: genotype-given-phenotype
posterior_genotype <- freq*background/K["Background"]
stopifnot(abs(sum(posterior_genotype)-1)<1e-12)
data.frame(genotype=names(freq),given_phenotype=unname(posterior_genotype))
c(phenotype_given_aa=background["aa"],aa_given_phenotype=posterior_genotype["aa"])
```

Although aa has high phenotype probability, many phenotype-positive individuals in this toy population have another genotype because those genotypes are much more common. $P(Y=1\mid aa)$ and $P(aa\mid Y=1)$ answer different questions.

This calculation presumes that all probabilities and the population sampling scheme are appropriate. A phenotype-ascertained sample has a different genotype distribution from a random population sample; its allele frequency cannot be treated as the population frequency without accounting for sampling.

## 6. Can phenotype prevalence identify q?

Only under full recessive penetrance, zero background probability and HWE does $q=\sqrt K$. If aa penetrance is known to be f with zero background, $q=\sqrt{K/f}$ instead. If F is nonzero, even the fully penetrant expression becomes $K=q^2+Fq(1-q)$.

```{r}
#| label: reverse-frequency-assumptions
observed_K <- 0.0004
penetrances <- c(1,0.8,0.5)
q_scenarios <- sqrt(observed_K/penetrances)
data.frame(supplied_aa_penetrance=penetrances,q=q_scenarios,
           check_K=penetrances*q_scenarios^2)
naive_q_background <- sqrt(K["Background"])
c(true_supplied_q=q,naive_sqrt_prevalence=naive_q_background)
```

Several combinations of q and penetrance produce the same K. The naive square-root inversion in the background example overestimates the supplied q. Unknown background probability or population structure adds further ambiguity. Observing prevalence alone does not identify all these quantities.

## Exercises

1. Recalculate heterozygote and aa frequencies at q=0.01 under HWE.
2. Compute genotype and phenotype probabilities for AA by Aa using the background model.
3. Give a penetrance vector for which increasing F lowers phenotype probability.
4. State the assumptions needed to infer q as the square root of prevalence.

::: {.callout-note collapse="true" title="Suggested solutions"}
```{r}
genotype_frequencies(0.01)
cross <- cross_probabilities(0,1)
cross
phenotype_probability(cross,background)
heterozygote_only <- c(AA=0,Aa=1,aa=0)
c(F0=phenotype_probability(genotype_frequencies(q,0),heterozygote_only),
  F01=phenotype_probability(genotype_frequencies(q,0.1),heterozygote_only))
```
At q=0.01, heterozygote probability is 0.0198 and aa probability is 0.0001. AA by Aa gives half AA and half Aa; its phenotype probability is 0.001 under the background model. If only heterozygotes express the phenotype, increasing F reduces its population probability. The square-root inversion requires HWE, a purely recessive fully penetrant single-locus phenotype, zero background cases and a prevalence measurement corresponding to the model's population and observation window.
:::

## Package connections and teaching extensions

The examples illustrate genotype-frequency expectations and conditioning relevant to `gpop`; they do not assume an exported medical-risk API. `gsim` can extend known crosses to explicit inheritance simulations, while fitness effects and selection policies require separate definitions. Penetrance is a phenotype probability and is not automatically a reproductive fitness weight.

Useful extensions include an explicitly specified age-dependent phenotype model, several alleles at one locus or a sampling model for uncertain allele frequencies. Each requires additional assumptions rather than substituting a new number into q squared.
